Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-08-28 thesis/data.tex

To provide an empirical benchmark for simulations in the next chapters,
we aim to understand the observed properties and density profiles of
Sculptor and Ursa Minor. In this section, we first compile past
measurements of each galaxy. We then describe the \citet{jensen+2024}
Bayesian methodology to select high probability member stars from
\emph{Gaia}. We show the derived density profiles are robust to changes
in assumptions. While Fornax is well represented by a 2D exponential
profile, Sculptor and Ursa Minor show an excess of stars relative to an
2D exponential in the outer regions, better described by a Plummer
profile. Future chapters assess possible explanations for these excess.

\section{\texorpdfstring{\emph{Gaia} data}{Gaia data}}\label{gaia-data}

Here, we briefly describe \citet{jensen+2024}'s (hereafter J+24)
membership estimation method. J+24 use a Bayesian framework
incorporating proper motion (PM), colour-magnitude diagram (CMD), and
spatial information to determine the probability that a given star
belongs to the satellite. By accounting for PM in particular, J+24
produces low contamination samples of candidate member stars out to
large distances from a dwarf galaxy. J+24 extends the algorithm
presented in \citet{MV2020a}; \citet{MV2020b} but additionally includes
an optional secondary, extended spatial component to find possible
members as far as \textasciitilde10 half-light radii \(R_h\) from
several dwarf galaxies. See also similar work by \citet{pace+li2019};
\citet{battaglia+2022}; \citet{pace+erkal+li2022}; \citet{qi+2022}.

J+24 define likelihoods \({\cal L}\) representing the probability
density that a star is consistent with either the MW stellar background
(\({\cal L}_{\rm bg}\)) or the satellite galaxy
(\({\cal L}_{\rm sat}\)). In either case, the likelihoods are the
product of a spatial, PM, and CMD term: \begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

Each likelihood is normalized over their respective 2D parameter space
for both the satellite. To control the relative frequency of member and
background stars, \(f_{\rm sat}\) representing the fraction of member
stars in the field. The total likelihood for any star in this model is
the sum of the satellite and background likelihoods, weighted by their
relative frequencies, \begin{equation}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
}\end{equation} The probability that any star belongs to the satellite
is then given by \begin{equation}{
P_{\rm sat} = 
\frac{f_{\rm sat}\,{\cal L}_{\rm sat}}{{\cal L}_{\rm tot}}
= \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
}\end{equation}

\section{Selected stars}\label{selected-stars}

@@

One limitation of the J+24 method is that it assumes a spatial form of
the dwarf galaxy, possibly limiting the detection of very distant stars
or features. Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection} also
illustrates the distribution of stars selected without a spatial
criterion. We define the \emph{CMD+PM} selection as
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
}\end{equation} with the likelihoods from J+24 as described above. These
stars are distributed similar to the fiducial (probable members) sample
but with a background uniform distribution across the entire field. This
illustrates the approximate background of stars which may be confused as
members. Additionally, since there is no clear spatial structure in the
\emph{CMD+PM} (or \emph{all}) sample outside several \(R_h\), it is
unlikely that there are additional faint, tidal features detectable with
\emph{Gaia}. Not shown here, we also try a variety of simpler, absolute
cuts and thresholds, finding no evidence of other structure in both
\emph{Gaia} and DELVE/UNIONS data.

Finally, we illustrate the location of RV-confirmed members from
Section~\ref{sec:rv_obs} in each panel. Because RV targets tend to be
brighter than a typical \emph{Gaia} candidate, RV members typically have
more precise PMs. However, the RV members fill about the same area on
the CMD down to the magnitude limit. Additionally, while the number of
stars with RV measurements decreases at large radii, there are still
confirmed RV members as far as \(\sim 10R_h\) for both galaxies. These
stars illustrate the extended profiles of each galaxy. Note that for an
exponential profile, 99.95\% of stars should be within \(6R_h\), so the
discovery of any stars beyond \(6R_h\) hints at a deviation from an
exponential profile. These \emph{extratidal} stars hint at something
tidal in nature, and our goal is to explore possible interpretations.

>>>>

An ongoing research question is whether local dwarf spheroidal galaxies
contain substructure such as stellar halos. Some systems appear to be
simple in structure, yet recent work has reported detection of stellar
halos in several classical dwarfs. To provide an empirical benchmark for
simulations in the next chapters, we examine the observed stellar
structure of Sculptor and Ursa Minor. We describe the
\citet{jensen+2024} Bayesian methodology to select high-probability
member stars from \emph{Gaia}. The derived density profiles are robust
to alternative selection criteria. While Fornax is well represented by
an exponential density profile, we find that Sculptor and Ursa Minor
exhibit an excess of stars in their outer regions relative to an
exponential, with a Plummer profile providing a better fit. In following
chapters, we consider possible physical explanations for these extended
profiles.

\section{\texorpdfstring{\emph{Gaia} sample
selection}{Gaia sample selection}}\label{gaia-sample-selection}

The detection of faint features in resolved stellar systems requires
careful consideration of whether a given star belongs to the system.
Without removing contamination from foreground/background features,
faint features may be lost in the noise or be of dubious association
with the system. Historically, background discrimination was performed
by the colour-magnitude diagram alone. Now with \emph{Gaia} data,
stellar parallax and proper motion areavailable as additional membership
indicators. Here, we use the \citet{jensen+2024}'s (hereafter J+24)
membership probabilities for \emph{Gaia} observations. J+24 use a
Bayesian framework incorporating proper motion (PM), colour-magnitude
diagram (CMD), and spatial information to determine the probability that
a given star belongs to the satellite or background. By accounting for
PM in particular, J+24 produces low contamination samples of candidate
member stars out to large distances from a dwarf galaxy. J+24 extends
the algorithm presented in \citet{MV2020a}; \citet{MV2020b} by
additionally includes a secondary, extended spatial component. J+24 are
able to find candidate members out to \textasciitilde10 half-light radii
(\(R_h\)). Similar recent work has also included \citet{pace+li2019};
\citet{battaglia+2022}; \citet{pace+erkal+li2022}; \citet{qi+2022}.

J+24 construct likelihoods, \({\cal L}\), representing the probability
density that a star is consistent with either the stellar
foreground/background, \({\cal L}_{\rm bg}\), or the satellite galaxy,
\({\cal L}_{\rm sat}\). In either case, the likelihoods are the product
of a spatial, PM, and CMD components, \begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

The spatial likelihood is specified either as a 1 or 2 component
elliptical exponential density profile (with the 2 component used when
shown to be detected). The proper motion likelihood specifies if a
star's proper motion is consistent with the dwarf galaxy's systemic
motion, accounting for observational uncertainties. Finally, the CMD
likelihood is based on the consistency of a star with theoretical
isochrones in \emph{Gaia} \(G\) versus \(G_{\rm BP}- G_{\rm RP}\) for
the galaxy. The background likelihood is a uniform distribution over the
field, and the background CMD and PM likelihoods are constructed
empirically from the stars in an annulus surrounding the dwarf galaxy.
Each likelihood is normalized over the respective parameter space.

The total likelihood is the sum of the background and satellite
likelihoods, weighted by the fraction of stars in the field belonging to
the satellite \(f_{\rm sat}\): \begin{equation}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
}\end{equation} The probability that any star belongs to the satellite
is then given by \begin{equation}{
P_{\rm sat} = 
\frac{f_{\rm sat}\,{\cal L}_{\rm sat}}{{\cal L}_{\rm tot}}
= \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
}\end{equation}

J+24 run MCMC simulations to solve for the proper motions,
\(f_{\rm sat}\), and structural properties of the second exponential
density profile (if included). The median parameter are then used to
calculate \(P_{\rm sat}\) which we use to select our samples of likely
member stars.

\section{Observed stellar
distributions}\label{observed-stellar-distributions}

@@

One limitation of the J+24 method is that it assumes a spatial form of
the dwarf galaxy, possibly limiting the detection of very distant stars
or features. Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection} also
illustrates the distribution of stars selected without a spatial
criterion. We define the \emph{CMD+PM} selection as
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}
}\end{equation} with the likelihoods from J+24 as described above. These
stars are distributed similar to the fiducial (probable members) sample
but with a background uniform distribution across the entire field. This
illustrates the approximate background of stars which may be confused as
members. Additionally, since there is no clear spatial structure in the
\emph{CMD+PM} (or \emph{all}) sample outside several \(R_h\), it is
unlikely that there are additional faint, tidal features detectable with
\emph{Gaia}.

Finally, we illustrate the location of distant, RV-confirmed members
from Section~\ref{sec:rv_obs} in each panel. These stars illustrate the
extended profiles of each galaxy. Note that for an exponential profile,
99.95\% of stars should be within \(6R_h\), so the discovery of any
stars beyond \(6R_h\) hints at a deviation from an exponential profile.
Distant member stars beg for an explanation as to their origin.

### 2025-08-28 thesis/sculptor.tex

Before running N-body simulations, we need to understand both the
typical halos which could host a sculptor like galaxy and the possible
orbits Sculptor.

Fig.~\ref{fig:scl_halos} presents approximate constraints based on

@@

\begin{table*}[t]
\centering
\caption[Derived Properties of Sculptor]{To derive total mass, we use the absolute magnitude from Muñoz et al. (2018) (see also Table \ref{tbl:scl_obs_props}) along with the stellar mass to light ratio (1.7 with 0.17 dex uncertainty) from Woo, Courteau, and Dekel (2008). Sculptor’s total stellar mass is then $M_\star \sim 3.1_{-1.0}^{1.6} \times 10^6\,\Mo$. From the stellar mass to dark matter halo’s characteristic velocity $v_{\rm circ}$ in Fattahi et al. (2018), we expect Sculptor to have $v_{\rm circ} \approx 31 \pm 3 \kms$. }
\label{tbl:scl_derived_props}
\begin{tabular}{lll}
\toprule
parameter & value & reference\\
\midrule
$L_\star$ & $1.8\pm0.2\times10^6\ L_\odot$ & \\
$M_\star$ & $3.1_{-1.0}^{+1.6} \times10^6\ {\rm M}_\odot$ & \\
$M_\star / L_\star$ & $1.7\times 10^{\pm 0.17}$ & Woo, Courteau, and Dekel (2008)\\
$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & \\
$c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
$v_{\rm circ, max}$ & $31\pm 3\,\kms$ & relation from Fattahi et al. (2018)\\
$r_{\rm circ, max}$ & $6 \pm 2$ kpc & $v_{\rm circ}$ with Ludlow et al. (2016) mass-concentration relation\\
$r_{\rm break, obs}$ & $25 \pm 5$ arcmin & (sestito+2024a?)\\
$t_{\rm break, obs}$ & $110\pm30$ Myr & $\sigma_v$, $r_{\rm break}$ with @\\
\bottomrule
\end{tabular}
\end{table*}

For the mass-to-light ratio, we use the values from
\citet{woo+courteau+dekel2008}. Note that we assume a scatter of 0.17
dex for the mass-to-light ratios as reported for mass uncertainties.
\citet{delosreyes+2024} also indicate that other methods for stellar
mass-to-light ratios result in similar scatters and systematic biases.
However these values are just approximate.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_initial_halos.pdf}}
\caption[Sculptor initial halos]{The suggested halos of Sculptor
compared to cosmological predictions.}\label{fig:scl_halos}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_xyzr_orbits.pdf}}
\caption[Sculptor Orbits]{The orbits of Sculptor in a static Milky Way
potential in galactocentric \(x\), \(y\), and \(z\) coordinates. The
Milky Way is at the centre with the disk lying in the \(x\)--\(y\)
plane. Our selected \texttt{smallperi} orbit is plotted in black and
light blue transparent orbits represent the past 5Gyr orbits sampled
over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The orbit of
sculptor is well-constrained in this potential and it is unlikely to
achieve a smaller pericentre than our selected orbit.
\the\textwidth}\label{fig:scl_orbits}
\end{figure}

To select an orbit with \textasciitilde{} maximum possible
observationally-consistent tidal forces, we take the median parameters
for orbits with a pericentre less than 2x the 3\(\sigma\) minimum
pericentre and derive the following orbit. The initial conditions are
from the

@@

\begin{itemize}
\tightlist
\item
  dark matter loss
\item
  little influence to stars
\item
  smallest pericentre does not help
\end{itemize}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_images.png}}
\caption[Sculptor simulation snapshots]{Images of the dark matter
evolution over a selection of past apocentres and the present day
position. Limits range from -150 to 150 kpc in the \(y\)-\(z\)
(approximately orbital) plane and the colourscale is logarithmic
spanning 5 orders of magnitude between the maximum and minimum values.
In this image, stars occupy only ever a few pixels so are not plotted.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.pdf}}
\caption[Sculptor Tidal Tracks]{The tidal tracks for the smallperi
orbit. Todo: add velocity dispersion plot to RHS}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sigma_v_time.pdf}}
\caption[Sculptor velocity dispersion evolution]{Evolution of stellar
velocity dispersion within 1 kpc for different Scl models. In all cases,
the evolution is mild. Note that binarity may reduce the inflate the
observed velocity dispersion by \textasciitilde{} 1 km/s, so the
conservative lower limit is around 8 km/s.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_smallperi_i_f.pdf}}
\caption[Sculptor initial and final density profiles]{Effects on
exponential initial stars. TODO: plot 2D sky proj. stars}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_plummer_i_f.pdf}}
\caption[Sculptor Plummer initial and final density profiles]{effects on
Plummer initial stars.}
\end{figure}

@@

@@

As discussed in \citet{battaglia+2022}, Sculptor's orbit is strongly
influenced by the presence of an LMC. Figure
Fig.~\ref{fig:scl_lmc_orbit_effect}, the addition of an LMC reduces
Scl's pericentre with the MW and implies that Scl may be on its first
infall with the MW, in contrast to the discussion above. Thus, the LMC
has a critical impact on the evolution of Scl. Additionally, Scl passes
rapidly but relatively close to the LMC in the past 100 Myr.

Finally, we also consider the influence of the large Milky Cloud on the
orbits and evolution of Scl. We adopt the \citet{vasiliev2024} multipole
approximation of an N-body simulation of the LMC and MW. Their initial
conditions are

\begin{itemize}
\tightlist
\item
  MW halo:
\item
  MW bulge (static):
\item
  MW disk (static):
\item
  LMC halo:
\end{itemize}

We now consider 3 orbits under the L3M11 potential model from
\citet{vasiliev2024}. We focus on this LMC potential as a lighter LMC or
MW should only reduce tidal impacts and the recent orbit of Scl is
minimally affected by the LMC potential.

@@

Smallperi

\begin{itemize}
\item
  ra = 15.0183 dec = -33.7186 distance = 84.0 pmra = 0.166 pmdec =
  -0.237 radial\_velocity = 111.2

  t\_i = -2.00 pericentre = 28.00 apocentre = 190.45 t last peri = -0.39
  x\_i = {[}12.79 168.76 87.33{]} v\_i = {[}0.37 -57.69 61.40{]}
\end{itemize}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_sim_images.pdf}}
\caption{Sculptor Simulation Snapshots with LMC}
\end{figure}

>>>>

@@

From the observed properties of Sculptor, we can infer reasonable
\(\Lambda\)CMD dark matter halo hosts. Table~\ref{tbl:scl_derived_props}
reports our inferred halo and kinematic properties of Sculptor. First,
taking the absolute magnitude from \citet{munoz+2018} with the
mass-to-light ratio from \citet{woo+courteau+dekel2008} (1.7 with
\(\sim\) 0.17 dex uncertainty), the total current stellar mass of
Sculptor is \(M_\star \sim 3.1 \times 10^6 \Mo\). Based on the stellar
mass-\(v_{\rm max}\) relation \citep[from][]{fattahi+2018}, see also
\ref{fig:SMHM}, the halo should have
\(v_{\rm max} \sim 31 \pm 3 \,\kms\). Finally, using the
\citet{ludlow+2016} \(z=0\) mass-concentration relation, this constraint
translates into a \(r_{\rm max} \sim 6 \pm 2\,{\rm kpc}\), or
alternatively \(M_{200} = 5\times10^9\) with \(c=13\).

Fig.~\ref{fig:scl_halos} illustrates these estimates visually. The
stellar-mass \(v_{\rm max}\) constraint translates to an estimate of
\(v_{\rm max}\) only (horizontal band). The mass-concentration relation
describes the relationship between \(v_{\rm max}\) and \(r_{\rm max}\),
an approximately diagonal linear band. And finally, we include a curved
line which illustrates where the initial velocity dispersion is
approximantly larger than the present-day observed dispersion
\(v_{\rm circ}(R_h) / \sqrt{3} \sim \sigma_v \gtrsim 9\,\kms\).

The intersection of the constraints on the halo leads us to select an
initial halo, \texttt{compact}, which is more compact than the mean but
with a observationally consistent initial velocity dispersion. We also
have the \texttt{light} halo which occupies the intersection between
3\(\sigma\) lowest mass and the velocity dispersion constraints, which
we come back to in Section~\ref{sec:scl_lmc}.

While there is some range in the choice of initial halo, reasonable
changes to the initial halo do not substantially affect the tidal
evolution for Sculptor. The velocity dispersion, in particular,
translates to an estimate of the mass within \(r_h\). As a result, halos
with the same velocity dispersion only differ in total mass, but the
density slope and circular velocity near the stellar component is
similar. These halos will have a similarly affected stellar component.

\begin{table*}[t]
\centering
\caption[Derived Properties of Sculptor]{To derive total mass, we use the absolute magnitude from Muñoz et al. (2018) (see also Table \ref{tbl:scl_obs_props}) along with the stellar mass to light ratio (1.7 with 0.17 dex uncertainty) from Woo, Courteau, and Dekel (2008). Sculptor’s total stellar mass is then $M_\star \sim 3.1_{-1.0}^{1.6} \times 10^6\,\Mo$. From the stellar mass to dark matter halo’s characteristic velocity $v_{\rm circ}$ in Fattahi et al. (2018), we expect Sculptor to have $v_{\rm circ} \approx 31 \pm 3 \kms$. }
\label{tbl:scl_derived_props}
\begin{tabular}{lll}
\toprule
parameter & value & reference\\
\midrule
$L_\star$ & $1.8\pm0.2\times10^6\ L_\odot$ & \\
$M_\star$ & $3.1_{-1.0}^{+1.6} \times10^6\ {\rm M}_\odot$ & \\
$M_\star / L_\star$ & $1.7\times 10^{\pm 0.17}$ & Woo, Courteau, and Dekel (2008)\\
$v_{\rm circ, max}$ & $31\pm 3\,\kms$ & relation from Fattahi et al. (2018)\\
$r_{\rm circ, max}$ & $6 \pm 2$ kpc & $v_{\rm circ}$ with Ludlow et al. (2016) mass-concentration relation\\
$M_{200}$ & $0.5 \pm 0.2\times10^{10}\ M_0$ & \\
$c_{\rm NFW}$ & $13.1_{-2.8}^{+3.6}$ & \\
$r_{\rm break, obs}$ & $25 \pm 5$ arcmin & (sestito+2024a?)\\
$t_{\rm break, obs}$ & $110\pm30$ Myr & $\sigma_v$, $r_{\rm break}$ with @\\
\bottomrule
\end{tabular}
\end{table*}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_initial_halos.pdf}}
\caption[Sculptor initial halos]{The suggested halos of Sculptor
compared to cosmological predictions. \textbf{TODO: simplify
\(\sigma_v\) calculation}.}\label{fig:scl_halos}
\end{figure}

@@

Next, we calculate the range of possible orbits of Sculptor in our
fiducial Milky Way potential. Fig.~\ref{fig:scl_orbits} illustrates
point particle orbits for 100 samples of Sculptor's observed kinematics
integrated backwards 5 Gyr in both galactocentric coordinate slices (x,
y, z) and in galactocentric radius with time. In terms of the x, y, z
planes, note that all sampled orbits of Sculptor have nearly the same
morphology---they each orbit mostly in the y-z plane, passing through
the same number of pericentres and maintaining similar orbits across
5Gyr. Sculptor's orbital history is relatively well-constrained.

We select an orbit with the \(\sim 3\sigma\) smallest pericentre among
all possible orbits. We achieve this by taking the median parameters of
all orbits with a pericentre less than the 0.0027th quantile pericentre,
yielding a pericentre of 43 kpc.\footnote{only slightly smaller than the
  0.00135th quantile pericentre} It is unlikely given the current
observations that Sculptor has a smaller pericentre. Since tidal effects
depend most strongly on the pericentre of an orbit, this orbit should
maximize the tidal force from the Milky Way.

We take the first apocentre after a lookback time of 10 Gyr, or at 9.43
Gyr, as the initial conditions for our model of Sculptor.
Table~\ref{tbl:scl_orbits} notes the initial conditions for our
simulation.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_xyzr_orbits.pdf}}
\caption[Sculptor Orbits]{The orbits of Sculptor in a static Milky Way
potential in galactocentric \(x\), \(y\), and \(z\) coordinates. The
Milky Way is at the centre with the disk lying in the \(x\)--\(y\)
plane. Our selected \texttt{smallperi} orbit is plotted in black and
light blue transparent orbits represent the past 5Gyr orbits sampled
over Sculptor observables in Table~\ref{tbl:scl_obs_props}. The orbit of
sculptor is well-constrained in this potential and it is unlikely to
achieve a smaller pericentre than our selected
orbit.}\label{fig:scl_orbits}
\end{figure}

@@

The Milky Way tides indeed affect Sculptor's dark matter halo.
Fig.~\ref{fig:scl_sim_images} shows projected dark matter distribution
of our N-body simulation for Sculptor in the \(y\)-\(z\) plane for 5
previous pericentres. This simulation uses the \texttt{compact} initial
halo on the \texttt{smallperi} orbit. The dark matter forms large
streams which orbit the Milky Way several times. However, note that the
central dark matter cusp is similar in appearance across all simulation
snapshots.

More quantitatively, the halo loses about 99\% of its initial dark
matter mass. This corresponds in a reduction from the initial
\(v_{\rm max} = 31\,\kms\) to \(v_{\rm max} = 22\,\kms\). However, the
stellar component of Sculptor is relatively unaffected.

While mass loss leads to adiabatic expansion of the stellar component,
no features of tidal disruption or disequilibrium are currently
observable. Fig.~\ref{fig:scl_smallperi_i_f} shows the final projected
density of stars on the sky, and the initial and final radially averaged
density profiles. No non-spherical density features are apparent, even
at 5 decades fainter in surface density than the central density. In
addition, the initial and final density profiles look identical up to
some scale, with a tidally-induced excess of stars only appearing
\(\sim 3\) orders of magnitude fainter than the faintest our density
profile measures.

The analytically motivated break and tidal radii corroborate a weak
tidal effect on stars. In both cases, these radii work out to be
\(\gtrsim 100\) arcminutes, outside where we can measure the density
profile. In particular, because we have chosen the most extreme
observationally permissible orbit, it is unlikely, for similar
\(\Lambda\)CDM motivated initial conditions, that these radii would ever
approach where the break in Sculptor's density profile appears.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_images.png}}
\caption[Sculptor simulation snapshots]{Images of the dark matter
evolution over a selection of past apocentres and the present day
position. Limits range from -150 to 150 kpc in the \(y\)-\(z\)
(approximately orbital) plane and the colourscale is logarithmic
spanning 5 orders of magnitude between the maximum and minimum values.
In this image, stars occupy only ever a few pixels so are not
plotted.}\label{fig:scl_sim_images}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_tidal_track.pdf}}
\caption[Sculptor Tidal Tracks]{The tidal tracks for the smallperi
orbit. Todo: add velocity dispersion plot to
RHS}\label{fig:scl_tidal_track}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sigma_v_time.pdf}}
\caption[Sculptor velocity dispersion evolution]{Evolution of stellar
velocity dispersion within 1 kpc for different Scl models. In all cases,
the evolution is mild. Note that binarity may reduce the inflate the
observed velocity dispersion by \textasciitilde{} 1 km/s, so the
conservative lower limit is around 8 km/s. \textbf{TODO}, add bound mass
with time. Maybe combine with tidal tracks and radius / time orbit.}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_smallperi_i_f.pdf}}
\caption[Sculptor initial and final density profiles]{Effects on
exponential initial stars. TODO: plot 2D sky proj.
stars}\label{fig:scl_smallperi_i_f}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_plummer_i_f.pdf}}
\caption[Sculptor Plummer initial and final density profiles]{effects on
Plummer initial stars.}\label{fig:scl_smallperi_plummer_i_f}
\end{figure}

@@

The Milky Way isn't the only galaxy in town. Becoming more apparent, the
infall of the LMC substantially affects the Milky Way system. The mass
of the LMC may be as high as 1/5 of the Milky Way mass, so including an
LMC in the Milky Way potential may change conclusions about properties
and orbits of the Milky Way satellite system.

@@

As discussed in \citet{battaglia+2022}, Sculptor's orbit is strongly
influenced by the presence of an LMC.

To explore the effects of the LMC on Sculptor, we use the publically
available L3M11 model from \citet{vasiliev2024}. This model stars with a
lighter Milky Way than our fiducial \citet{EP2020}, \citet{mcmillan2011}
like model. The L3M11 potential is an evolving multipole approximation
of an N-body simulation with both a live MW and LMC dark matter halo.
The MW intital conditions were an NFW with \(r_s=16.5\,\)kpc and mass
\(M_{\rm 100}= 11\times10^{11}\Mo\), and the LMC was a NFW halo with
\(r_s=11.7\) and \(M_{100} = 2.76 \times 10^{11} \Mo\). This model had a
previous LMC pericentre at about 6 Gyr ago.

Fig.~\ref{fig:scl_lmc_orbit_effect} displays the effect of including an
LMC in the potential. The green samples are in the initial MW-only
potential in the \texttt{L3M11} model, and the orange samples are
integrated in the evolving MW and LMC \texttt{L3M11} model. The past 1
Gyr is similar in both cases, but the orbits diverge significantly
afterwards. The recent passage of Sculptor with the LMC around 0.1 Gyr
ago allows for Sculptor to begin as far as 300 kpc from the Milky Way
centre. The evolving potential also adds significant long term
variability in the possible orbits of Sculptor. Sculptor, however, is
orbiting in the opposite direction of the LMC so is likely not
associated with the LMC system.

Given the large uncertainties of the LMC model, we conservatively double
all of the observational parameters of Sculptor. This has a similar
effect to including LMC mass and orbital uncertainties but is
considerably simpler. From this larger range of orbits, we once again
select the orbit with the median final observables of all orbits with
pericentres less than the 0.0027th quantile. This orbit, the
\texttt{smallperilmc} orbit is plotted in black and is only integrated
up to 2 Gyr ago, to isolate recent tidal effects.

@@

@@

Unlike the previous mdoel, this model only has one pericentre with the
Milky Way. Fig.~\ref{fig:scl_lmc_sim_images} shows snapshots of Sculptor
over the past 2 Gyr while marking the position of the LMC. With only one
pericentre, and a larger one than the \texttt{smallperi} orbit,
Sculptor's dark matter is substantially less disrupted. And while, based
on the tidal tensor values, the LMC induces a greater instantaneous
tidal effect than the Milky Way, Sculptor's dark matter component does
not show strong effects due to the LMC.

Finally, Fig.~\ref{fig:scl_lmc_i_f} shows the projected on-sky final
stellar distribution and the initial and final stellar density profile
for this model with exponential stars. The stellar component does not
change at all. While the break radius set by the time since the last LMC
pericentre would agree with the location of the break in the observed
density profile, no stellar effect would be observable.

The LMC flyby encounter is an approximately impulsive encounter, in
contrast with more adiabadic mass loss due to the Milky Way. Impulsive
encounters tend to inject energy into the stellar and dark matter
distribution, and can initially cause the galaxy to contract. In
addition, the tidal force is required to be far larger than for slower,
adiabadic mass loss, because the galaxy experiences this tidal field for
far less time. Even in models where Sculptor passes through the the LMC,
tidal tails do not form immediately after this encounter.

A final note is that this model only considers the recent tidal effects
.As illustrated in the range of possible orbits in
Fig.~\ref{fig:scl_lmc_orbits_effect}, there is a chance that Sculptor
experienced an extremely small pericentre with the Milky Way. This
pericentre has the potential to substantially rearrange the stellar
component, drive large tidal mass loss, and create a Plummer-like
stellar density profile. However, this encounter is highly dependent on
the choice of the MW-LMC potential model, and may not occur at all. Long
time integration in dynamic potentials amplifies uncertainties in the
inputs and may not be reliable. \textbf{TODO: I would love to get this
model to work\ldots{}}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_lmc_sim_images.pdf}}
\caption{Sculptor Simulation Snapshots with
LMC}\label{fig:scl_lmc_sim_images}
\end{figure}
